---
title: "Phoenix Crime Trends"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: yeti
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 12, 
                      fig.height = 10)

rm(list = ls())

```

<br><br>

## Has crime in Phoenix increased during the COVID-19 pandemic?  

I was wondering whether there has been an increase in crime in Phoenix since March 2020. So, I went to the city's [open data site](https://www.phoenixopendata.com) and pulled the [crime data](https://www.phoenixopendata.com/dataset/crime-data/resource/0ce3411a-2fc6-4302-a33f-167f68608a20) which is updated every day.  

<br><br>

### Getting the data.  

Let's pull the most recent data from the site.

```{r, echo = TRUE, eval = TRUE}

# set the url where the data are located.
url <- "https://www.phoenixopendata.com/dataset/cc08aace-9ca9-467f-b6c1-f0879ab1a358/resource/0ce3411a-2fc6-4302-a33f-167f68608a20/download/crimestat.csv"

# pull in the csv file.
crime.data <- read.csv(url, as.is = TRUE, header = TRUE)

# take a look at the data.
head(crime.data)  
tail(crime.data)
dim(crime.data)

```  

The data are reported as UCR crime classifications and have geographic information (block address, zip) as well as the date and time of the incident. The `dim()` command returns 315,386 reports.

<br><br>

### Cleaning the data.  

Now that the data are in the workspace, let's clean up the date and the crime categories to make plotting them fairly easy. For this step, we will use the  `lubridate()` package.

```{r, echo = TRUE, eval = TRUE}

# Load the library.
library(lubridate)

# Check the class of the data variable.
#crime.data$OCCURRED.ON

# Check for format of the occurred on date variable.
crime.data$OCCURRED.ON[1] # this is month, day, year as well as the time (hours and minutes).
mdy_hm(crime.data$OCCURRED.ON[1])

# create an object that is the time and the category.
dates.crimes <- data.frame(
  date = as.Date(mdy_hm(crime.data$OCCURRED.ON)),
  crime = crime.data$UCR.CRIME.CATEGORY,
  days.since = unclass(as.Date(mdy_hm(crime.data$OCCURRED.ON))) 
  )

table(is.na(dates.crimes$date)) # 420 are missing dates.
table(is.na(dates.crimes$crime)) # none missing type of crime.

# Let's drop those cases missing dates.
dates.crimes <- dates.crimes[is.na(dates.crimes$date) == FALSE,]
dim(crime.data)[1] - dim(dates.crimes)[1] # the number of cases you had to drop due to missing.

# The days.since object represents the number of days since 1/1/1970.
# We got this using the unclass() function. I like working with the numbers.
head(dates.crimes$days.since)

```

<br><br>

### Plot it!

Now, we want to plot the data where the months are the x-axis and the frequency of the crimes are the y-axis. We have to set up some plot parameters first to get things looking nice.

```{r, echo = TRUE, eval = TRUE}

# Set up the values we want to plot.
y.1 <- as.vector(table(dates.crimes$days.since - dates.crimes$days.since[1]))
x.1 <- c(seq(min(dates.crimes$days.since - dates.crimes$days.since[1]),max(dates.crimes$days.since - dates.crimes$days.since[1]),length.out = length(y.1)))

# unclass(as.Date("2020-03-01")) - dates.crimes$days.since[1] # this is March 1st 2020.
march.1.2020 <- (unclass(as.Date("2020-03-01")) - dates.crimes$days.since[1])
march.1.2019 <- (unclass(as.Date("2019-03-01")) - dates.crimes$days.since[1])
march.1.2018 <- (unclass(as.Date("2018-03-01")) - dates.crimes$days.since[1])
march.1.2017 <- (unclass(as.Date("2017-03-01")) - dates.crimes$days.since[1])
march.1.2016 <- (unclass(as.Date("2016-03-01")) - dates.crimes$days.since[1])
marches      <- cbind(march.1.2019,march.1.2018,march.1.2017,march.1.2016)

### Set plot basics and axes.
plot.new()
plot.window(xlim = c(0, max(x.1)), ylim = c(min(y.1), max(y.1)+10))   

### Plot the values.  
points(x = x.1, y = y.1, col = "black", pch = 16, cex = 0.75)
lines(x = x.1, y = y.1, col = "grey75", pch = 16, cex = 0.75)

# Add a segment to shown the median.
#segments(0,median(y.1),max(x.1),median(y.1), lwd = 3, lty = 3, col = "black")

# Add a line and some text for March 2020.
segments(march.1.2020, min(y.1), march.1.2020, 270,lwd = 2, lty = 3, col = "red")
text(march.1.2020, 275,"March 1, 2020", col="red", pos=1, cex=1.1)

# Add a line and some text for March in the prior years.
for(i in 1:length(marches)){
  segments(marches[i], min(y.1), marches[i], 270,lwd = 1, lty = 2, col = "black")
}
text(marches[1], 275,"March 1, 2019", col="black", pos=1, cex=0.8)
text(marches[2], 275,"March 1, 2018", col="black", pos=1, cex=0.8)
text(marches[3], 275,"March 1, 2017", col="black", pos=1, cex=0.8)
text(marches[4], 275,"March 1, 2016", col="black", pos=1, cex=0.8)

# Add a lowess line to see the trend.
lines(lowess(x.1, y.1), type = "l", col= "black", lwd = 2)

# Add a title.
title("Reports of Crime for Phoenix (Nov. 2015 - Sept. 2020)\n[lowess shown in horizontal black line]")

```

<br><br>

### Conclusions?  

This is just a quick look, but it shows that crime went down after the pandemic hit in Phoenix. There is a brief uptick though. So, perhaps we should take a closer look at the 2020 data.
